Management induced changes of soil organic carbon on global croplands

### 2022-05-25 SOCbudgetPaper/methods.Rmd
updated Paper version

---
output:
  pdf_document: default
  html_document: default
---
# Methods

In Sect. \@ref(sec:carbonbudget) we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol. 4 [@calvo_buendia_ipcc_2019]. We additionally describe how we configured and extended the steady-state method (for model code see [@karstens_mrsoil_2020]. In Sect. \@ref(sec:tier1) we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019]. 
Section \@ref(sec:agrimanagement) provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices [for model code see @bodirsky_mrcommons_2020]. In Sect. \@ref(sec:scenarios) we define the scenarios used to complement our historic model results.

## SOC stocks and stock changes following the steady-state method {#sec:carbonbudget}

Following the Tier 2 steady-state approach of the refinement of the IPCC guidelines vol. 4 [@calvo_buendia_ipcc_2019]; referred to as \textit{steady-state method} in the following), we estimate soil organic carbon ($SOC$) stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current $SOC$ state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period: 
(1) We calculate annual land-use type-specific steady states and decay rates for $SOC$ stocks (Sect. \@ref(sec:steadystates)); 
(2) we account for land conversion by transferring $SOC$ from and to natural vegetation (Sect. \@ref(sec:carbontransfer)),
(3) we estimate $SOC$ stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \@ref(sec:totalsoc)). 
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \@ref(sec:initsoc)).

@@

```{r captionLUtrans, echo=FALSE}
captionLUtrans <- "Scheme of land-use transition representation for a simplified case.."
```

```{r LUtrans, out.width = "16cm", fig.cap = captionLUtrans, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images//LUtransition.png")
```

We calculate $SOC$ stocks based on the area shares of land-use types ($lu$) within our grid cells ($i$). If land is converted from one land-use type $lu=\{crop,natveg\}$ into the other $!lu=\{natveg,crop\}$, a respective share of the $SOC$ is reallocated within our budget. As we do not distinguish between newly converted and old cropland (see \@ref(fig:lutransitions)), we update the average SOC stock per area within this modeling step. We account for land conversion at the beginning of each time step $t$ by calculating a preliminary stock $SOC_{t^*}$ via

@@

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lu} \overbrace{\sum_{sub} SOC_{i,t,sub,lu}}^{\text{$SOC_{i,t,lu}$ -- land-use type specific SOC stock within cell}}}_{\text{$SOC_{i,t}$ -- total SOC stock within cell}}.
(\#eq:totalstock)
\end{equation}

According to the IPCC guidelines $SOC$ changes can be expressed as the difference of two consecutive years [see Eq. 5.0A in @calvo_buendia_ipcc_2019]. This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we defined the absolute and relative SOC changes in relation to a potential natural state $SOC^{\mathrm{pnv}}$ under the same climatic conditions in grid cell $i$ at time $t$, that is based on the natural vegetation $SOC$ calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes $\Delta SOC$ and relative changes $F^{\mathrm{SCF}}$ are thus given by

@@

We initialize our SOC sub-pools using a three-step approach, since input data availability is limited for climate and litter (from 1901) as well as for agricultural management data (from 1965):

Firstly, to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we account for land-use change from 1510. We repeat climate and subsequently also litter input data from 1901-1930 to simulate a constant climate for this first initializing period. Agricultural management data is hold constant at the level of 1965. In 1510 we assume all SOC pools to be in natural steady-state, implicating that all land-use change emissions prior to that time, are redistributed to 1510. Due to this long first spin-up period we suppose that the effect of this late accounting and appertaining delayed legacy fluxes can be neglected in 1901. This approaches follows others studies looking on effects of land-use change and management (cite LPJmL (which?)).

Secondly, to account for the impact of a changing climate, we account not only for land-use change, but also for historical climate and consequently natural litter inputs to the soil, from 1901 to 1965 still considering constant agricultural input data.

Thirdly, we run the model for additional 10 years with historic dynamic input data and start analyzing results from 1975 onward. This is in line with IPCC method suggestion to have a 5-20 year spin-up period.

Irrigation areas are part of the water effect associated with the climate data set and therefore dynamic from 1901 on, whereas data on no-tillage areas is only available after 1974.

<!-- backup
```{=latex}
 \begin{table*}[h]
 \caption{Setup of spin-up and simulation phase}
 \begin{tabular}{l l l}
 \tophline
 \multirow{3}{*}{1901--1964} 
   & Constant management spin-up phase with
   & historic land-use change (LUH2v2), climate (CRU4p2) and litter inputs (LPJmL4)
   & constant management data from 1965 (manure and residue inputs, tillage and irrigation areas)\\
                      \cline{1-4}
 \multirow{3}{*}{1965--1974}
   & Dynamic management spin-up phase with
   & historic land-use change (LUH2v2), climate (CRU4p2) and litter inputs (LPJmL4)
   & historic management data (manure and residue inputs, tillage and irrigation areas)
                      \cline{1-4}
 \multirow{3 }{*}{1975--2010}                     
   & Simulation phase 
   & historic land-use change (LUH2v2), climate (CRU4p2) and litter inputs (LPJmL4)
   & historic management data (manure and residue inputs, tillage and irrigation areas)              
 \bottomhline
 \end{tabular}
 \belowtable{}
 (\#tab:datasourcedecay)
 \end{table*}
``` -->
## SOC stocks and stock changes following Tier 1 {#sec:tier1}

Additionally to the steady-state method [@calvo_buendia_ipcc_2019] and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019]. Here, stocks are calculated via stock change factors ($F^{\mathrm{SCF}}$) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of $F^{\mathrm{SCF}}$ are differentiated by different crops, management and input systems (here summarized under $m$) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, estimates of $F^{\mathrm{SCF}}$ vary for different climatic zones ($c$) specified by the IPCC (see Fig. \@ref(fig:CLIMzone)). The actual $SOC$ stocks are thus calculated based on a given reference stock $SOC^{\mathrm{ref}}$ by

@@

Land-use patterns are based on the Land-Use Harmonization 2 [@hurtt_harmonization_2020] data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups [see FAO2LUH2MAG_croptypes.csv in @bodirsky_mrcommons_2020], applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data [@faostat_faostat_2016]. By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level. 
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL [@schaphoff_lpjml4_2018] as well as half-degree cropland patterns (see Sect. \@ref(sec:landuse)). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historic FAOSTAT production [@faostat_faostat_2016]. By using physical cropland areas in combination with harvested areas, we account for multiple cropping systems as well as for fallow land.

Crop residue production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and key aspects are explained here given its central role for soil carbon modeling. Starting from crop yield estimates ($Y$) and respective harvested crop area ($CA$), we estimate above-ground ($AGR$) and below-ground ($BGR$) residual biomass using crop group ($cg$) specific ratios for above-ground residues to harvested biomass ($r^{\mathrm{ag,prod}}_{cg}$ in $\tfrac{\unit{tC}}{\unit{tC}}$), above-ground residues to harvested area ($r^{\mathrm{ag,area}}_{cg}$ in $\tfrac{\unit{tC}}{\unit{ha}}$) and below-ground residues to above-ground biomass ($r^{\mathrm{bg}}_{cg}$ in $\tfrac{\unit{tC}}{\unit{tC}}$) as follows

\begin{equation}
\begin{aligned}
AGR_{i,t,cg} & = CA_{i,t,cg} \cdot \left( Y_{i,t,cg} \cdot  r^{\mathrm{ag,prod}}_{cg}+  r^{\mathrm{ag,area}}_{cg} \right)
\qquad & \textrm{and} \\
BGR_{i,t,cg} & = (CP_{i,t,cg} + AGR_{i,t,cg}) \cdot r^{\mathrm{bg}}_{cg} \qquad                                            & \forall\quad cg, i, t.
(\#eq:resbiomass)
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the above-ground residue calculations into a production ($r^{\mathrm{ag,prod}}$) and an area dependent ($r^{\mathrm{ag,area}}$) fraction [@hergoualch_n2o_in_ipcc_2019]. Deviating from Bodirsky et al. [-@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

A country-specific fixed share of the $AGR$ is assumed to be burned on field depending on the per-capita income of the country. Following Smil [-@smil_nitrogen_1999] we assume a burn share of $25\%$ for low-income countries according to World Bank definitions ($<\,1000\,\tfrac{USD}{yr\,cap}$), 15% for high-income  ($>\,10000\,\tfrac{USD}{yr\,cap}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80–90% of the carbon in the crop residues burned in the fields is lost within the combustion process [@eggleston_ipcc_2006].

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al. [-@bodirsky_mrcommons_2020]. Residue feed demand for five different livestock groups is based on country-specific feed baskets [see @weindl_livestock_2017], that differentiate between the residue groups and take available $AGR$ biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5% and a fuel-share of 10% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining $AGR$ as well as all $BGR$ are expected to be left on the field. We limit high residue return rates to at most $10\unit{tC\,ha}^{-1}$ in order to correct for outliers.

To transform dry matter estimates into carbon and nitrogen, we compiled crop-group and plant-part specific carbon resp. nitrogen to dry matter (c/dm, n/dm) ratios (see \@ref(tab:c2dm)).

@@

Manure especially from ruminants is often excreted at pastures and rangelands, but due to the intensification of livestock systems 	at the present day a lot of the manure has to be stored and can be applied on croplands. We assume that manure application happens at its excretion place, so that the livestock distribution is the driving factor of the spatial pattern of manuring.

To disaggregate country level FAOSTAT livestock production data to half-degree resolution, we use the following rule-based assumptions, drawing from the approach of Robinson et al. [-@robinson_mapping_2014] and applying feed basket assumptions based on a revised methodology from Weindl et al. [-@weindl_livestock_2017]. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
Due to great feed demand of ruminants, we assume that ruminant livestock is located where the production of feed occurs to minimize transport of feed. We distinguish between grazed pasture, which is converted into livestock products in extensive systems, and primary-crop feed stuff, which we consider to be consumed in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to distinguish between intensive and extensive production systems. For low-income countries, we assume only extensive production systems. We locate them according to the share of built-up areas based on the assumption that these animals are held in subsistence or small-holder farming systems with a high labor-per-animal ratio. Intensive production associated with high-income countries, is distributed within a country using the share of primary-crop production, assuming that feed availability is the most determining factor for livestock location. For middle-income countries we split the livestock production into extensive and intensive systems based on the per-capita income.

Manure production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets [@weindl_livestock_2017], assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see Calvo Buendia et al. [-@calvo_buendia_ipcc_2019]. 
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates (see calcClossConfinement.R in Bodirsky et al. [-@bodirsky_mrcommons_2020] for more details) depending on the different systems and the associated nitrogen loss rates as specified in Bodirsky et al. [-@bodirsky_n2o_2012]. We limit high application shares at $10\unit{tC\,ha}^{-1}$ to correct for outliers.

@@

The LUH2v2 [@hurtt_harmonization_2020] data set provides irrigated fractions for their cropland subcategories. We sum up irrigation area shares for all crop groups within a grid cell, and calculate the water effect coefficient $wat$ on decay rates using these shares to compute the weighted mean between rainfed and irrigated $wat$ factors. As a result $wat$ is the same for all crop groups within a grid cell. Furthermore, we suppose the irrigation effect to be present for all 12 months of a year in a grid cell including irrigated areas, since we do not have consistent crop group specific growing periods available. This will lead to an overestimation of the irrigation effect. We expect, however, water limitations to be a minor problem during the off-season in temperature limited cropping regions, causing our assumption to not dramatically overestimate the moisture effects. In tropical, water-limited cropping areas, irrigated growing periods might even span over the whole year.

@@

In order to derive a spatial distribution of the three different tillage types specified by the IPCC --- full tillage, reduced tillage and no tillage ---, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage per default. Furthermore, we assume no tillage in cropland cells specified as no tillage cell based on the historic global gridded tillage data set from Porwollik et al. [-@porwollik_generating_2018]. This data set is extended to the period of 1975--2010 by combining country-level data on areas under conservation agriculture from FAO [-@fao_aquastat_2016] and half-degree resolution physical crop areas from Hurtt et al. [-@hurtt_harmonization_2020], applying the methodology of Prowollik et al. [-@porwollik_generating_2018] to identify potential no-tillage grid cells.

## Scenario definitions {#sec:scenarios}

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization and parameter choices, we perform a set of scenario runs. In the following sections we outline name and idea of these scenarios [for technical implementation see @karstens_mrsoil_2020].

### Management scenarios {#sec:scen_management}

>>>>

# Methods

In Sect. \@ref(sec:carbonbudget) we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail in the refinement of the IPCC guidelines vol. 4 Chapter 5 on "Cropland" [@ogle_cropland_in_ipcc_2019]. We additionally describe how we configured and extended the Tier 2 modeling approach [for model code see @karstens_mrsoil_2020]. In Sect. \@ref(sec:tier1) we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019]. 
Section \@ref(sec:agrimanagement) provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices [for model code see @bodirsky_mrcommons_2020]. In Sect. \@ref(sec:scenarios) we define the management scenarios used to analyze the role of different management aspects in historical cropland SOC dynamics.

## SOC stocks and stock changes following the Tier 2 modeling approach {#sec:carbonbudget}

Following the Tier 2 modeling approach of the refinement of the IPCC guidelines vol. 4 Chapter 5 on "Cropland" [@ogle_cropland_in_ipcc_2019]; referred to as \textit{Tier 2 modeling approach} in the following), we estimate $SOC$ stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current $SOC$ state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period: 
(1) We calculate annual land-use type-specific steady states and decay rates for $SOC$ stocks (Sect. \@ref(sec:steadystates)); 
(2) we account for land conversion by transferring $SOC$ from and to natural vegetation (Sect. \@ref(sec:carbontransfer)),
(3) we estimate $SOC$ stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \@ref(sec:totalsoc)). 
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \@ref(sec:initsoc)).

@@

```{r captionLUtrans, echo=FALSE, include=FALSE}
captionLUtrans <- "Scheme of land-use transition representation. Given an initial land-use pattern (as in this example 2 ha land under natural vegetation and 1 ha of cropland), there are separate $SOC$ stocks for natural vegetation and cropland. While in this example we assume $SOC$ under natural vegetation to be in steady state, the cropland $SOC$ stock approaches its steady state without having reached it yet (a). Upon cropland expansion (in this example half of the natural vegetation is cleared to be used as cropland), $SOC$ stocks on cropland increase due to a transfer of land from natural vegetation (b). Explicitly representing newly converted cropland and existing cropland to account for SOC dynamics (c) leads to the same weighted mean value as averaging $SOC$ stocks (d), due to the linearity of Eq. \\ref{eq:SOCstate} and cropland-age independent decay rates (see Eq. \\ref{eq:decayrates})."
```

```{r LUtrans, out.width = "8cm", fig.cap = captionLUtrans, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images/LUtransitions.png")
```

We calculate $SOC$ stocks based on the area shares of land-use types $lu$ within our grid cells $i$. If land is converted from one land-use type $lu=\{crop,natveg\}$ into the other $!lu=\{natveg,crop\}$, a respective share of the $SOC$ is reallocated within our budget. We do not distinguish between newly converted and existing cropland, but can work with the average carbon content as the relative decay of $SOC$ is proportional to the $SOC$ stock (see \@ref(fig:LUtrans)). We account for land conversion at the beginning of each time step $t$ by calculating a preliminary stock $SOC_{t^*}$ via

@@

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lu} \overbrace{\sum_{sub} SOC_{i,t,sub,lu}}^{\text{$SOC_{i,t,lu}$ -- land-use type specific $SOC$ stock within cell}}}_{\text{$SOC_{i,t}$ -- total $SOC$ stock within cell}}.
(\#eq:totalstock)
\end{equation}

According to the IPCC guidelines $SOC$ changes can be expressed as the difference of two consecutive years [see Eq. 5.0A in @ogle_cropland_in_ipcc_2019]. This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we define the absolute and relative $SOC$ changes in relation to a potential natural state $SOC^{\mathrm{pnv}}$ under the same climatic conditions in grid cell $i$ at time $t$, that is based on the natural vegetation $SOC$ calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes $\Delta SOC$ and relative changes $F^{\mathrm{SCF}}$ are thus given by

@@

The initialization of $SOC$ pools is very important and has to include the proper accounting for the land-use history, as many CO~2~ emissions from agricultural soils are caused by historical land-use change (LUC) and the slow decline of $SOC$ under crop cultivation, before it reaches a new equilibrium.
We initialize our $SOC$ sub-pools using a three-step approach, since input data availability is limited for climate and litter estimates (starting only in 1901) as well as for agricultural management data (starting only in 1965):

Firstly, in order to account for the impacts of legacy fluxes from land-use changes long before the time horizon of interest, we consider land-use change from 1510 onwards. In 1510, we assume all $SOC$ pools to be in natural steady-state, implying that all land-use change prior to that time occurs in 1510. We assume that by 1901, all cropland converted in 1510 has reached its new steady state, so that it is not necessary to explicitly account for even older land conversion. Model inputs for 1901--1930 for climate and natural vegetation litterfall are repeated for 1510--1900 to mimic constant climate conditions for this first initializing period. Similarly, agricultural management data are held constant at the level of 1965 until 1965.  This approach follows others studies looking on effects of land-use change and management [e.g. @schaphoff_lpjml4_2018-1; @herzfeld_soc_2021].

Secondly, with the availability of transient climate data after 1901, we account not only for land-use change, but also for historical climate change and consequently natural litter inputs to the soil from 1901 to 1965 still considering constant agricultural input data, which are not available prior to 1965.

Thirdly, we run the model for 10 years from 1965 to 1975 with historical dynamic data on agricultural management and start analyzing results from 1975 onward. This is in line with the IPCC guidelines vol. 4 method suggestion to have a 5-20 year spin-up period [@ogle_cropland_in_ipcc_2019].

With transient climate considered after 1901, decay rates $k_{sub}$ become dynamic in time. As the decay rates are also affected by irrigation and tillage (see Sect. \@ref(sec:steadystates)), we also account for transient changes in irrigated areas after 1901. Data on no-tillage practices are only available after 1974 and we assume conventional tillage on all cropland prior to 1975.

## SOC stocks and stock changes following Tier 1 {#sec:tier1}

Additionally to the Tier 2 modeling approach [@ogle_cropland_in_ipcc_2019] and the detailed analysis of management data coming with it, $SOC$ changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines [@eggleston_ipcc_2006; @calvo_buendia_ipcc_2019]. Here, stocks are calculated via stock change factors ($F^{\mathrm{SCF}}$) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of $F^{\mathrm{SCF}}$ are differentiated by crops, management and input systems (here summarized under $m$) reflecting different dynamics under changed in- and outflows without explicitly tracking these flows. Moreover, estimates of $F^{\mathrm{SCF}}$ vary for different climatic zones ($c$) specified by the IPCC (see Fig. \@ref(fig:CLIMzone)). The actual $SOC$ stocks are thus calculated based on a given reference stock $SOC^{\mathrm{ref}}$ by

@@

Land-use patterns are based on the Land-Use Harmonization 2 [@hurtt_harmonization_2020] data set (short LUH2), which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area (given as total land area in million ha) of the five different cropland subcategories (c3ann: C3 annual crops, c3per: C3 perennial crops, c4ann: C4 annual crops, c4per: C4 perennial crops, c3nfx: C3 nitrogen-fixing crops) of LUH2 into our 17 crop groups [see Table "FAO2LUH2MAG_croptypes.csv" in @karstens_model_2020], applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data [@faostat_faostat_2016]. By calculating country-specific multicropping factors $MCF$ using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level. 
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

Crop production patterns are compiled crop group specific using half-degree yield data from LPJmL [@schaphoff_lpjml4_2018] as well as half-degree cropland patterns (see Sect. \@ref(sec:landuse)). We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production [@faostat_faostat_2016]. Note

Crop residue production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and key aspects are explained as they play a central role in soil carbon modeling. Starting from crop yield estimates $Y$ and respective physical crop area $CA$, we estimate total above-ground $AGR$ and below-ground $BGR$ residue biomass (in tonnes) using crop group $cg$ specific ratios for above-ground residues to harvested biomass $r^{\mathrm{ag,prod}}_{cg}$ in $(tDM\,ha^{-1})(tDM\,ha^{-1})^{-1}$ above-ground residues to harvested area $r^{\mathrm{ag,area}}_{cg}$ in $\unit{tDM\,\unit{ha}^{-1}}$ and below-ground residues to above-ground biomass $r^{\mathrm{bg}}_{cg}$ in $\unit{tDM\,\unit{tDM}^{-1}}$ as follows

\begin{equation}
\begin{aligned}
AGR_{i,t,cg} & = CA_{i,t,cg} \cdot \left( Y_{i,t,cg} \cdot  r^{\mathrm{ag,prod}}_{cg} + MCF_{i,t} \cdot r^{\mathrm{ag,area}}_{cg} \right)
\qquad & \textrm{and} \\
BGR_{i,t,cg} & = (CA_{i,t,cg} \cdot Y_{i,t,cg} + AGR_{i,t,cg}) \cdot r^{\mathrm{bg}}_{cg} \quad.
(\#eq:resbiomass)
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the above-ground residue calculations into a yield-dependent slope ($r^{\mathrm{ag,prod}}$) and a positive intercept ($r^{\mathrm{ag,area}}$) fraction [@hergoualch_n2o_in_ipcc_2019]. Residues biomass therefore increases under-proportionally with rising yields, reflecting a shifting harvest index of higher-yielding breeds. 
Deviating from Bodirsky et al. [-@bodirsky_n2o_2012] we use harvested instead of physical crop area (denoted in Eq. \@ref(eq:resbiomass) by MCF described in Sect. \@ref(sec:landuse)) to account for increased residue biomass due to multiple cropping (multiple harvests with each lower yields) and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

A country-specific fixed share of the $AGR$ is assumed to be burned on field depending on the per-capita income of the country. Following Smil [-@smil_nitrogen_1999] we assume a burn share of 25% for low-income countries according to World Bank definitions ($<\,10000\,\unit{USD}\,\unit{yr}^{-1}\,\unit{cap}^{-1}$), 15% for high-income  ($>\,10000\,\unit{USD}\,\unit{yr}^{-1}\,\unit{cap}^{-1}$) and linearly interpolate shares for all middle-income countries depending on their per-capita income for the periods before 1995. After 1995 we estimate a linear decline of burn shares to 10% for low-income countries and 0% for high-income countries till 2025 to account for recent increases in air pollution regulation. The estimated trends show good agreement with fire-satellite-image derived estimates by the Global Fire Database [@vanderWerf_fire_2017]. Depending on the crop group, 80–90% of the carbon in the crop residues burned in the fields is lost within the combustion process [@eggleston_ipcc_2006].

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al. [-@bodirsky_mrcommons_2020]). Residue feed demand for five different livestock groups is based on country-specific feed baskets [see @weindl_livestock_2017], that differentiate between the residue groups and take available $AGR$ biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5% and a fuel-share of 10% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining $AGR$ as well as all $BGR$ are expected to be left on the field. We limit high residue return rates to at most $10\,\unit{tC\,ha}^{-1}$ in order to correct for outliers.

To transform dry matter estimates into carbon and nitrogen, we compiled crop-group and plant-part specific carbon and nitrogen to dry matter ratios (see Table \@ref(tab:c2dm)).

@@

Manure especially from ruminants is often excreted at pastures and rangelands, but due to the intensification of livestock systems a lot of the manure has to be stored and can be applied on cropland. We assume that manure is applied in close proximity to its excretion, so that the distribution of livestock is the driving factor for the spatial pattern of manure application.

To disaggregate country level FAOSTAT livestock production data to half-degree resolution, we use the following rule-based assumptions, drawing from the approach of Robinson et al. [-@robinson_mapping_2014] and applying feed basket assumptions based on a revised methodology from Weindl et al. [-@weindl_livestock_2017]. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
Due to high feed demand of ruminants, we assume that ruminant livestock is located where the production of feed occurs to minimize transport of feed. We distinguish between grazed pasture, which is converted into livestock products in extensive systems, and primary-crop feed stuff, which we consider to be consumed in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to distinguish between intensive and extensive production systems. For low-income countries, we assume only extensive production systems. We locate them according to the share of built-up areas based on the assumption that these animals are held in subsistence or small-holder farming systems with a high labor-per-animal ratio. Intensive production associated with high-income countries, is distributed within a country using the share of primary-crop production, assuming that feed availability is the most determining factor for livestock location. For middle-income countries we split the livestock production into extensive and intensive systems based on the per-capita income.

Manure production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets [@weindl_livestock_2017], assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios, which range from 10 for poultry up to 19 for beef cattle (for full detail see Calvo Buendia et al. [-@ogle_cropland_in_ipcc_2019]. 
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates (see calcClossConfinement.R in Bodirsky et al. [-@bodirsky_mrcommons_2020] for more details) depending on the different systems and the associated nitrogen loss rates as specified in Bodirsky et al. [-@bodirsky_n2o_2012]. We limit high application shares at $10\,\unit{tC\,ha}^{-1}$ to correct for outliers, that can occur due to inconsistencies between FAO production and 0.5 degree land-use data.

@@

The LUH2v2 [@hurtt_harmonization_2020] data set provides irrigated fractions for its cropland subcategories. We sum up irrigation area shares for all crop groups within a grid cell, and calculate the water effect coefficient $wat$ on decay rates using these shares to compute the weighted mean between rainfed and irrigated $wat$ factors. As a result $wat$ is the same for all crop groups within a grid cell. Furthermore, we suppose the irrigation effect to be present for all 12 months of a year, since we do not have consistent crop group specific growing periods available. This will lead to an overestimation of the irrigation effect. We expect, however, water limitations to be a minor problem during the off-season in temperature limited cropping regions, causing our assumption to not dramatically overestimate the moisture effects. In tropical, water-limited cropping areas, irrigated growing periods might even span over the whole year.

@@

In order to derive a spatial distribution of the three different tillage types specified by the IPCC --- full tillage, reduced tillage and no tillage ---, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage per default. Furthermore, we assume no tillage in cropland cells specified as no tillage cell based on the historical global gridded tillage data set from Porwollik et al. [-@porwollik_generating_2019]. This data set is extended to the period of 1975--2010 by combining country-level data on areas under conservation agriculture from FAO [-@fao_aquastat_2016] and half-degree resolution physical crop areas from Hurtt et al. [-@hurtt_harmonization_2020], applying the methodology of Prowollik et al. [-@porwollik_generating_2019] to identify potential no-tillage grid cells.

## Management scenario definitions {#sec:scenarios}
